---
title: "Non-linear dimensionality reduction. (Assignment)"
author: "Louis Van Langendonck & Antonin Rosa & Albert Martín"
date: "`r format(Sys.time(), '%d/%b/%Y')`"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(princurve)
```

## Part A: Principal Curves

### 1. Choosing the smoothing parameter in Principal Curves

Generate dataset as instructed:

```{r}
t <- seq(-1.5*pi,1.5*pi,l=100)
R<- 1
n<-75
sd.eps <- .15

set.seed(1)
y <- R*sign(t) - R*sign(t)*cos(t/R)
x <- -R*sin(t/R)
z <- (y/(2*R))^2
rt <- sort(runif(n)*3*pi - 1.5*pi)
eps <- rnorm(n)*sd.eps
ry <- R*sign(rt) - (R+eps)*sign(rt)*cos(rt/R)
rx <- -(R+eps)*sin(rt/R)
rz <- (ry/(2*R))^2 + runif(n,min=-2*sd.eps,max=2*sd.eps)
XYZ <- cbind(rx,ry,rz)


require(plot3D)
lines3D(x,y,z,colvar = NULL, 
         phi = 20, theta = 60, r =sqrt(3), d =3, scale=FALSE,
         col=2,lwd=4,as=1,
         xlim=range(rx),ylim=range(ry),zlim=range(rz))
points3D(rx,ry,rz,col=4,pch=19,cex=.6,add=TRUE)
```


#### a) LOOCV for df:

To compute the Leave One Out Cross-Validation, we split the samples in five folds, and for each of them fit a Principal Curve to the other four folds, and compute the distance of the remaining fold to the fitted PC. The sum of this distance across all five folds is the score for each value of df, so the one that leads to the lowest sum of squared distances is computed in the end, which ends up being 8.

```{r}
#Generate random permutation for the data, which represents the folds
set.seed(1)
nfolds<-3
foldSize<-n/nfolds
perm<-sample(1:n,n)


dfVals<-seq(2,8,by=1)
dfDist<-rep(0,length(dfVals))
for(dfIdx in 1:length(dfVals))
{
  df<-dfVals[dfIdx]
  for(fold in 1:nfolds){
    #Get each fold train and test splits
    testPerm<-perm[((fold-1)*foldSize+1):(fold*foldSize)]
    foldTrain=XYZ[-testPerm,]
    foldTest=XYZ[testPerm,]
    #Fit PC to train split
    foldPC<-principal_curve(foldTrain,df=df)
    #Get result from test split
    foldDist<-project_to_curve(foldTest,foldPC$s)$dist
    dfDist[dfIdx]<-dfDist[dfIdx]+foldDist
  }
}
min(dfDist)
df<-dfVals[which.min(dfDist)]
pc<-principal_curve(XYZ,df=df)
df
```

#### b) Visual representation of the curve

We can plot the Principal Curve along with the real line and the sample points by using the returned projection of the data to the line as guide points. The plot shows how the calculated curve follows the same trend as the real line and sample data points, but it does not match the former perfectly. This is due to the error introduced artificially when generating the sample, which shifts the line along the direction of the random spread.

We also notice the irregularities at the edges, which are a byproduct of a high df. This leads the PC to become less smooth as it tends to jump from data point to data point, which decreases ability to generalize.

```{r}
lines3D(x,y,z,colvar = NULL, 
         phi = 20, theta = 60, r =sqrt(3), d =3, scale=FALSE,
         col=2,lwd=4,as=1,
         xlim=range(rx),ylim=range(ry),zlim=range(rz))
lines3D(pc$s[,1],pc$s[,2],pc$s[,3],colvar = NULL, 
         phi = 20, theta = 60, r =sqrt(3), d =3, scale=FALSE,
         col=3,lwd=4,as=1,
         xlim=range(rx),ylim=range(ry),zlim=range(rz),add=TRUE)
points3D(rx,ry,rz,col=4,pch=19,cex=.6,add=TRUE)
legend("topright", legend = c("Source line", "Principal Curve","Sample points"), 
       col = c(2, 3, 4), lty = c(1, 1, 0), pch = c(NA, NA, 19), 
       lwd = c(4, 4, NA), 
       title = "Line", cex = 0.8)
```

#### c) LOOCV for df=50

To compute the Leave One Out Cross-Validation, we split the samples in five folds, and for each of them fit a Principal Curve to the other four folds, and compute the distance of the remaining fold to the line just as in ex.1 but with df=50.

Before computing this value, we believe it will be worse as before, from just increasing df from 6 to 8 it already showed clear signs of overfitting, with 50 it should be much more extreme.

```{r}
set.seed(1)
nfolds<-5
foldSize<-n/nfolds
perm<-sample(1:n,n)
distSum<-0
for(fold in 1:nfolds){
  testPerm<-perm[((fold-1)*foldSize+1):(fold*foldSize)]
  foldTrain=XYZ[-testPerm,]
  foldTest=XYZ[testPerm,]
  foldPC<-principal_curve(foldTrain,df=50)
  foldDist<-project_to_curve(foldTest,foldPC$s)$dist
  distSum<-distSum+foldDist
}
distSum
```

As expected, the new curve is even less smooth. In fact, it pretty much just jumps from point to point. While the LOOCV score is better than for df=8, the newly fitted PC is clearly less generalizable than before, which shows us that simply using the LOOCV score is not enough to determine which value for df is the ideal one.

We believe this behavior is due to the fact that the error is measured as the squared distance from the point to the line, and this value does not change much regardless of whether the fitted line is exactly like the real one, or that line is jumping between two close points. Even if it is visually clear that the new line does not adjust to the reality, the distances from the point to the real line are not that far off from the distances to the fitted line that jumps around other close data points.

```{r}
pc<-principal_curve(XYZ,df=50)
lines3D(x,y,z,colvar = NULL, 
         phi = 20, theta = 60, r =sqrt(3), d =3, scale=FALSE,
         col=2,lwd=4,as=1,
         xlim=range(rx),ylim=range(ry),zlim=range(rz))
lines3D(pc$s[,1],pc$s[,2],pc$s[,3],colvar = NULL, 
         phi = 20, theta = 60, r =sqrt(3), d =3, scale=FALSE,
         col=3,lwd=4,as=1,
         xlim=range(rx),ylim=range(ry),zlim=range(rz),add=TRUE)
points3D(rx,ry,rz,col=4,pch=19,cex=.6,add=TRUE)
legend("topright", legend = c("Source line", "Principal Curve","Sample points"), 
       col = c(2, 3, 4), lty = c(1, 1, 0), pch = c(NA, NA, 19), 
       lwd = c(4, 4, NA), 
       title = "Line", cex = 0.8)
```


## Part B : Local MDS, ISOMAP and t-SNE.

We load in the ZIP data and the necessary packages. Every row in this data represents a drawing of the number (we will only select numbers zero) and each column represents a pixel-value. Given that the images are 16x16, there are 256 columns, hence high-dimensional data. We will therefore try 3 dimensionality-reduction methods: Local-MDS, ISOMAP and t-SNE.

```{r}

if (!require("smacofx", quietly = TRUE, warn.conflicts = FALSE)) {
  install.packages("smacofx", repos = "http://R-Forge.R-project.org", INSTALL_opts = "--no-test-load")
}

library("smacofx")

if (!require(stops, quietly=TRUE, warn.conflicts=FALSE)){
  install.packages("stops", repos="http://R-Forge.R-project.org",INSTALL_opts="--no-test-load")
}

library(stops)

plot.zip <- function(x,use.first=FALSE,...){
  x<-as.numeric(x)
  if (use.first){
    x.mat <- matrix(x,16,16)
  }else{
    x.mat <- matrix(x[-1],16,16)
  }
  image(1:16,1:16,x.mat[,16:1],
        col=gray(seq(1,0,l=12)),...)
  #col=gray(seq(1,0,l=2)))
}

# Load the entire training data set
data <- read.table("zip.train", sep=" ")

# Separate the target labels and feature vectors
labels <- data$V1
features <- data[, -1]

# Filter only the ZEROs (label 0)
zero_indices <- which(labels == 0)
zero_data <- features[zero_indices,]
```

### 2. Local MDS for ZERO digits.

### a).

Using the proposed parameters of $k = 5$ and $\tau = 0.05$, a local MDS run is executed. Note that these typically takes a lot of time to run. The resulting two-dimensional representation of the data is plotted. 

```{r}
n <- dim(zero_data)[1]
k <- 5
tau <- .05
q<-2 # 2-dim config
conf0 <- stats::cmdscale(dist(zero_data), k=q)


lmds_result <- lmds(as.matrix(dist(zero_data)), init=conf0, ndim=2, k=5, tau=0.05, itmax = 1000)

ylower_limit <- min(lmds_result$conf[, 2]) - 0.1
yupper_limit <- max(lmds_result$conf[, 2]) + 0.1

xlower_limit <- min(lmds_result$conf[, 1]) - 0.1
xupper_limit <- max(lmds_result$conf[, 1]) + 0.1


plot(lmds_result$conf, type = "n", xlab = "Dimension 1", ylab = "Dimension 2", main = "LMDS Scatterplot", ylim = c(ylower_limit, yupper_limit), xlim=c(xlower_limit, xupper_limit))
points(lmds_result$conf, pch = 20)
```

### b).

As a strategy to select interesting points, for each of the dimensions, the point corresponding to the smallest value, the 1st quartile value, the median, the 3rd quartile value and the largest value of that dimension are selected. This adds up in total to 10 points that indeed cover the variation well (see red dots). We make a function of this method to select points to reuse it easily later.

```{r}
select_interesting_points <- function(config_points) {
    res <- c(which(config_points[,1]==min(config_points[,1])))
    res <- c(res,which(config_points[,1]==quantile(config_points[,1], 0.25, type=1)))
    res <- c(res,which(config_points[,1]==quantile(config_points[,1], 0.5, type=1)))
    res <- c(res,which(config_points[,1]==quantile(config_points[,1], 0.75, type=1)))
    res <- c(res,which(config_points[,1]==max(config_points[,1])))
    
    res <- c(res,which(config_points[,2]==min(config_points[,2])))
    res <- c(res,which(config_points[,2]==quantile(config_points[,2], 0.25, type=1)))
    res <- c(res,which(config_points[,2]==quantile(config_points[,2], 0.5, type=1)))
    res <- c(res,which(config_points[,2]==quantile(config_points[,2], 0.75, type=1)))
    res <- c(res,which(config_points[,2]==max(config_points[,2])))
    return(res)
}
```


```{r}
interesting_points <- select_interesting_points(lmds_result$conf)
selected_points <- lmds_result$conf[interesting_points, ]

plot(lmds_result$conf, xlab = "Dimension 1", ylab = "Dimension 2", main = "LMDS Scatterplot")
points(selected_points, pch = 20, col = "red")
text(selected_points, labels = interesting_points, pos = 3, cex = 0.7)
```

Plotting the corresponding drawings og the first five points (representing increasing first dimension), we can see this dimension seems to be proportional to increasing width and increasing diameter of the zero. 

```{r}
par(mfrow=c(2,3))
for (i in 1:5) {
  cat("Image for Point ", interesting_points[i], ":\n")
  plot.zip(zero_data[interesting_points[i], ])
  title(interesting_points[i])
}
```

From the drawing, the second dimension is a bit less easy to characterize. However, it seems to be correlated with increasingly atypical top parts of the drawing (like open parts, poorly drawn zeros, noise,...) or maybe inversely related with the vertical component of the zeros. 

```{r}
par(mfrow=c(2,3))
for (i in 6:10) {
  cat("Image for Point ", interesting_points[i], ":\n")
  plot.zip(zero_data[interesting_points[i], ])
  title(interesting_points[i])
}
```
### c).

Now the LC-meta-criteria is used to tune parameters. The LCMC function that takes as input the distance matrices of the high-dimensional and low-dimensional representations is copied from the example in class. This function calculates the LC-meta-criteria value for which the configuration giving the maximum value is chosen as optimal.

```{r}
LCMC <- function(D1,D2,Kp){
  D1 <- as.matrix(D1)
  D2 <- as.matrix(D2)
  n <- dim(D1)[1]
  N.Kp.i <- numeric(n)
  for (i in 1:n){
    N1.i <- sort.int(D1[i,],index.return = TRUE)$ix[1:Kp]
    N2.i <- sort.int(D2[i,],index.return = TRUE)$ix[1:Kp]
    N.Kp.i[i] <- length(intersect(N1.i, N2.i))
  }
  N.Kp<-mean(N.Kp.i)
  M.Kp.adj <- N.Kp/Kp - Kp/(n-1)
  
  return(list(N.Kp.i=N.Kp.i, M.Kp.adj=M.Kp.adj))
}
```

Now all combinations of the candidate values of k and tau are made and the LCMC calculated for there local MDS result. The configuration with the highest LCMC score is kept. We see that that is the case for k = 5 and tau = 1.

```{r}
D1 <- dist(zero_data)

k_values <- c(5, 10, 50)
tau_values <- c(0.1, 0.5, 1)

best_k <- NULL
best_tau <- NULL
best_lcmc <- -Inf  

best_configuration <- NULL

for (k in k_values) {
  for (tau in tau_values) {
    lmds_result <- lmds(as.matrix(dist(zero_data)), ndim = 2, k = k, tau = tau, itmax=1000)
    D2 <- dist(lmds_result$conf)
    lcmc <- LCMC(D1,D2,10)$M.Kp.adj
    
    if (lcmc > best_lcmc) {
      best_lcmc <- lcmc
      best_k <- k
      best_tau <- tau
      best_configuration <- lmds_result$conf
    }
  }
}

LC_lmds <- best_lcmc
cat("Best Configuration - k:", best_k, "tau:", best_tau, "LCMC:", best_lcmc, "\n")
```

The same strategy as before is maintained to select points covering the distribution of the scatterplots.

```{r}
interesting_points <- select_interesting_points(best_configuration)
plot(best_configuration, asp=1,xlab = "Dimension 1", ylab = "Dimension 2", main = "Optimal Low-Dimensional Configuration")
points(best_configuration[interesting_points,1],best_configuration[interesting_points,2],col = "red", pch=19)
text(best_configuration[interesting_points,1],best_configuration[interesting_points,2],labels = interesting_points, pos = 3, cex = 0.7, col = "red")
```

Plotting the points representing increasing value of dimension 1, we see that it seems to be proportionate to an increase in diameter of the zero and a decrease in width of the stroke.

```{r}
par(mfrow=c(2,3))
for (i in 1:5) {
  cat("Image for Point ", interesting_points[i], ":\n")
  plot.zip(zero_data[interesting_points[i], ])
  title(interesting_points[i])
}
```

Plotting the points representing the increase in dimension 2, we find a less obvious relation (as expected maybe from the scatterplot where most variance is in the first dimension). An interpretation could be that it is proportional to a decrease in the vertical component (vertical diameter) of the zero scribble.

```{r}
par(mfrow=c(2,3))
for (i in 6:10) {
  cat("Image for Point ", interesting_points[i], ":\n")
  plot.zip(zero_data[interesting_points[i], ])
  title(interesting_points[i])
}
```

### 3 ISOMAP for ZERO digits.

### a).

```{r}
library(vegan)

isomap_result <- isomap(as.matrix(dist(zero_data)), ndim = 2, k = 5)

isomap_coordinates <- isomap_result$points

plot(isomap_coordinates, type = "n", xlab = "Dimension 1", ylab = "Dimension 2", main = "Isomap 2-Dimensional Configuration")
points(isomap_coordinates, pch = 20, col = "blue")
```

### b).

```{r}
interesting_points <- select_interesting_points(isomap_coordinates)
plot(isomap_coordinates, asp=1)
points(isomap_coordinates[interesting_points,1],isomap_coordinates[interesting_points,2],col = "red", pch=19)
text(isomap_coordinates[interesting_points,1],isomap_coordinates[interesting_points,2],labels = interesting_points, pos = 3, cex = 0.7, col = "red")
```

From the images of the zeros, the first dimension can be interpreted as proportional to the diameter of the zero.

```{r}
par(mfrow=c(2,3))
for (i in 1:5) {
  cat("Image for Point ", interesting_points[i], ":\n")
  plot.zip(zero_data[interesting_points[i], ])
  title(interesting_points[i])
}
```

From the images of the zeros, the second dimension can be interpreted as proportional to the vertical component of the zero or inversely to the width of the stroke.

```{r}
par(mfrow=c(2,3))
for (i in 6:10) {
  cat("Image for Point ", interesting_points[i], ":\n")
  plot.zip(zero_data[interesting_points[i], ])
  title(interesting_points[i])
}
```

### c).

Similar to in part c) of the last question, the LCMC function is used to see for which low-dimensional configuration the LC-meta-criteria scores the highest. Three proposed values of perplexity are tested. A plot of the interesting points is then made. 

```{r}
D1 <- dist(zero_data)

perplexity <- c(5, 10, 50)
LC <- c(0, 0, 0)

for (i in 1:length(perplexity)) {
  isomap_result <- isomap(as.matrix(dist(zero_data)), ndim = 2, k = perplexity[i])
  D2 <- dist(isomap_result$points)
  LC[i] <- LCMC(D1, D2, perplexity[i])$M.Kp.adj
}

best_k <- perplexity[which.max(LC)]
LC_isomap <- max(LC)

cat("Accordig to the LC-criteria, the best value of k is:", best_k, "\n")

best_isomap <- isomap(as.matrix(dist(zero_data)), ndim = 2, k = best_k)

interesting_points <- select_interesting_points(best_isomap$points)

plot(best_isomap$points, asp=1)
points(best_isomap$points[interesting_points,1],best_isomap$points[interesting_points,2],col = "red", pch=19)
text(best_isomap$points[interesting_points,1],best_isomap$points[interesting_points,2],labels = interesting_points, pos = 3, cex = 0.7, col = "red")
```

From the images of the zeros, the first dimension can be interpreted as proportional to the diameter of the zero or the width of the zero.

```{r}
par(mfrow=c(2,3))
for (i in 1:5) {
  cat("Image for Point ", interesting_points[i], ":\n")
  plot.zip(zero_data[interesting_points[i], ])
  title(interesting_points[i])
}
```

From the images of the zeros, the second dimension can be interpreted as inversely proportional to stroke width.

```{r}
par(mfrow=c(2,3))
for (i in 6:10) {
  cat("Image for Point ", interesting_points[i], ":\n")
  plot.zip(zero_data[interesting_points[i], ])
  title(interesting_points[i])
}
```

### 4 t-SNE for ZERO digits.

### a)

Run rtsne and do scatterplot.

```{r}
#install.packages("Rtsne")
library(Rtsne)

zero_data_rsne <- zero_data[,-ncol(zero_data)] #Remove last column of missing values as the Rtsne package can't handle that.

set.seed(1)
tsne_out <- Rtsne(zero_data_rsne,dims=2,pca=FALSE,perplexity=40,theta=0.0) # Run TSNE

# Scatterplot of 2D tsne representation
plot(tsne_out$Y, asp=1)
```

### b). 

The same interesting points function as in previous exercises is used and the corresponding scatterplot shown.

```{r}
interesting_points <- select_interesting_points(tsne_out$Y)

plot(tsne_out$Y, asp=1)
points(tsne_out$Y[interesting_points,1],tsne_out$Y[interesting_points,2],col = "red", pch=19)
text(tsne_out$Y[interesting_points,1],tsne_out$Y[interesting_points,2],labels = interesting_points, pos = 3, cex = 0.7, col = "red")
```

Looking at the 5 images that represent increasing value of the first dimension it can be concluded that the first dimension seems to be increasing with the width of the drawing.

```{r}
par(mfrow=c(2,3))
for (i in 1:5) {
  cat("Image for Point ", interesting_points[i], ":\n")
  plot.zip(zero_data[interesting_points[i], ])
  title(interesting_points[i])
}
```

Looking at the 5 images that represent increasing value of the second dimension it can be concluded that the second dimension seems to be inversly related to the size of the diameter of the zero and positively related with the stroke width.

```{r}
par(mfrow=c(2,3))
for (i in 6:10) {
  cat("Image for Point ", interesting_points[i], ":\n")
  plot.zip(zero_data[interesting_points[i], ])
  title(interesting_points[i])
}
```
### c).

The three proposed values of perplexity are set at 10, 20 and 40. A loop is ran in which the rtsne is executed for each of these perplexity values and the corresponding Local Continuity meta criteria value calculated. From these results we can see that the highest value is found for perplexity 20.


```{r}
D1 <- dist(zero_data_rsne)

perplexity <- c(10,20,40)
LC <- c(0,0,0)

for (i in 1:length(perplexity)){
    Rtsne.k <- Rtsne(D1, perplexity=perplexity[i], dims=2,
                          theta=0, pca=FALSE, max_iter = 1000)
    D2 <- dist(Rtsne.k$Y)
    LC[i] <- LCMC(D1,D2,10)$M.Kp.adj
    #print(c(i,j,LC[i,j]))
}

best_perplexity <- perplexity[which.max(LC)]
LC_tsne <- max(LC)

best_Rtsne <- Rtsne(D1, perplexity=best_perplexity, dims=2,
                          theta=0, pca=FALSE, max_iter = 1000)

interesting_points <- select_interesting_points(best_Rtsne$Y)
plot(best_Rtsne$Y, asp=1)
points(best_Rtsne$Y[interesting_points,1],best_Rtsne$Y[interesting_points,2],col = "red", pch=19)
text(best_Rtsne$Y[interesting_points,1],best_Rtsne$Y[interesting_points,2],labels = interesting_points, pos = 3, cex = 0.7, col = "red")
```

Looking at the 5 images that represent increasing value of the first dimension it can be concluded that the first dimension seems to be positively related to the size of the diameter of the zero.

```{r}
par(mfrow=c(2,3))
for (i in 1:5) {
  cat("Image for Point ", interesting_points[i])
  plot.zip(zero_data[interesting_points[i], ])
  title(interesting_points[i])
}
```

Looking at the 5 images that represent increasing value of the second dimension it can be concluded that the second dimension seems to be inversly related to the width of the drawing.

```{r}
par(mfrow=c(2,3))
for (i in 6:10) {
  cat("Image for Point ", interesting_points[i], ":\n")
  plot.zip(zero_data[interesting_points[i], ])
  title(interesting_points[i])
}
```

### 5. Compare Local MDS, ISOMAP and t-SNE for ZERO digits

### a).

The pairs function is used to see if dimensions of different methods might align. We find that this clearly is the case, however in some cases more than the others. 

Most strikingly, the first dimension shows a clear relation:
- All methods are directly proportional to each other, which indicates these methods find a similar most informative first representation dimension. 

The second dimension also shows some proportionality, however, the correlation is less clear: 
- Isomap and t-sne are kind of inversely proportional. 
- t-sne and lmds are tend to be more or less directly proportional. 
- Isomap and Lmds show a more clear negative correlation.

Finally, looking at agreements between the dimensions:
- Some minor positive correlations between tsne_dim2 and the first dimensions of all methods are found. This is the case as the 2-dimensional representation of tsne already shows some positive correlation and these methods all agree in that first dimension, hence these correlation show up again comparing to these other methods.

```{r}
df.methods <- data.frame(best_Rtsne$Y[,1],best_isomap$points[,1],
                         best_configuration[,1],best_Rtsne$Y[,2],
                         best_isomap$points[,2],best_configuration[,2])
colnames(df.methods) <- c("Rtsne_dim1","Isomap_dim1",
                          "Lmds_dim1","Rtsne_dim2",
                          "Isomap_dim2","Lmds_dim2")
pairs(df.methods)
```

### b).

According to the local continuity meta criteria, the t-SNE gives the best results, hence representing the high-dimensional continuity very well in its lower-dimensional representation. Given that it seemed to be finding this representation the quickest of all three methods, it might be concluded its the strongest technique of the three.

```{r}
options <- c("Local MDS", "Isomap", "t-SNE")
print(paste0("Method having the highest value for local continuity meta criteria is: ", options[which.max(c(LC_lmds, LC_isomap, LC_tsne))]))
```

