---
title: "GAMS fits for hirsutism data"
author: "Louis Van Langendonck & Antonin Rosa & Albert Martín"
date: "`r format(Sys.time(), '%d/%b/%Y')`"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

### GAMs for hirsutism data

First model : 

```{r}
gam_model1 <- gam(FGm12 ~ s(FGm0) + s(SysPres) + s(DiaPres) + s(weight) + s(height) + s(Treatment, k = 4), data = data)
summary(gam_model1)
vis.gam(gam_model1)
gam.check(gam_model1)
```

If we look at the summary only two parameters have a p-value inferior to 0.1 : FGm0, Treatment and SysPres.

Then, the gam check suggest the relation with height, weight and syspres is linear.

Second model :

```{r}
gam_model2 <- gam(FGm12 ~ s(FGm0) + SysPres + s(DiaPres) + weight + height + s(Treatment, k = 3), data = data)
summary(gam_model2)
vis.gam(gam_model2)
gam.check(gam_model2)
```

In looking the summary for the next model we should try to remove DiaPres en weight.

Third model : 

```{r}
gam_model3 <- gam(FGm12 ~ s(FGm0) + s(SysPres) + s(height) + s(Treatment, k=4), data = data)
summary(gam_model3)
vis.gam(gam_model3)
gam.check(gam_model3)
```

If wee look at the p-value of gam.check, we should try with SysPres, FGm0 and Treatment in linear parameter.

Fourth model : 

```{r}
gam_model4 <- gam(FGm12 ~ FGm0 + SysPres + s(height) + Treatment, data = data)
summary(gam_model4)
vis.gam(gam_model4)
gam.check(gam_model4)
```

All the p-values in the summary are under 0.05 however the p-value of the gam.check can suggest us to try with height in linear parameter.

Fifth model : 

```{r}
gam_model5 <- gam(FGm12 ~ FGm0 + SysPres + height + Treatment, data = data)
summary(gam_model5)
vis.gam(gam_model5)
gam.check(gam_model5)
```

This model is bad because two parameters have p-value superior to 0.1.

ANOVA Results : 

```{r}
gam_model5 <- gam(FGm12 ~ FGm0 + SysPres + Treatment, data = data)

gam_model6 <- gam(FGm12 ~ SysPres + Treatment, data = data)

anova_result_4_5 <- anova(gam_model4, gam_model5, test = "Chisq")
print(anova_result_4_5)

anova_result_5_6 <- anova(gam_model5, gam_model6, test = "Chisq")
print(anova_result_5_6)
```

According to the ANOVA, these three models appear equivalent as the p-values are below 0.05.
