---
title: "GAM fits for hirsutism data"
author: "Louis Van Langendonck & Antonin Rosa & Albert Martín"
date: "`r format(Sys.time(), '%d/%b/%Y')`"
output: html_document
---

Loading in mgcv package for the gam function and loading in the Hirsutism dataset. 

```{r}
library(mgcv)

hirs <- read.table("hirsutism.dat",header=T, sep="\t",fill=TRUE)
```

For preprocessing we take the following steps: 
- We set the treatment levels as factor. We remove the two variables not used for analysis: FGm3 and FGm6m. 
- We convert the height and FGm12 variables from string to numeric. 
- Lastly, we omit the rows containing NA values (9 rows), because if we don't, models that use only variables not containing NAs will be using more data than other which makes comparing them difficult. 

```{r}
hirs$Treatment <- as.factor(hirs$Treatment)
data = hirs[,!(names(hirs) %in% c('FGm3','FGm6'))]
data$height <- as.numeric(data$height)
data$FGm12 <- as.numeric(data$FGm12)
data <- na.omit(data)
```

For data exploration, we plot all variables against eachother to get an initial idea of how they relate to each-other. Some initial take-aways seem to be possible collinearity between the two pressure variables and the weight and height variables. Also note that only 91 datapoints are available in the dataset, which means that our models might easily be overly complex / overfitting.

```{r}
plot(data)
nrow(data)
```
As a baseline, the first model we start from is a glm model using all the explanatory variables available and look at its summary.

```{r}
gam0 <- gam(FGm12 ~ FGm0 + Treatment + SysPres + DiaPres + weight + height, data = data)

summary(gam0)
```

We notice low deviance and unsignificant variable contributions (the pressure, weight and height variables)

Let's now try a variable where smooth terms are applied to each predictor (Generalized Additive Model) except for treatment as it is a factor (so semi-parametric). Interaction between the factor and numericals will be looked into later. Compare the model to the GLM using an anova test (ANOVA tests whether the more complex model is significantly better at capturing the data than the simpler model or not).

```{r}
gam1 <- gam(FGm12 ~ s(FGm0) + Treatment + s(SysPres) + s(DiaPres) + s(weight) + s(height), data = data)

summary(gam1)
anova(gam0, gam1, test='F')
```
We see that the more complex GAM model performs significantly better than the GLM. First we notice that Syspress, weight and height don't need smoothing as there edf is 1. This can be confirmed by the following vis-plots where only DiaPress shows serious curvature. Combined linear variables give these flat surfaces.

```{r}
vis.gam(gam1, view=c("weight","height"),
        theta = 0, phi = 10, r = sqrt(3), d = 1)

vis.gam(gam1, view=c("SysPres","height"),
        theta = 0, phi = 30, r = sqrt(3), d = 1)

vis.gam(gam1, view=c("DiaPres","height"),
        theta = 0, phi = 30, r = sqrt(3), d = 1)
```

We try making these predictors back linear and see with the Anova test if this model is preferred.

```{r}
gam1.0 <- gam(FGm12 ~ s(FGm0) + Treatment + SysPres + s(DiaPres) + weight + height, data = data)
summary(gam1.0)
anova(gam1, gam1.0, test='F')
```

The answer to this is yes. model1.0 is the current best model.

However, we note that there are terms that are not significant in prediction. we remove these one-by-one (based on descending p-value) until all variables are significant.

```{r}
gam1.1 <- gam(FGm12 ~ s(FGm0) + Treatment + SysPres + weight + height, data = data)
summary(gam1.1)
gam1.2 <- gam(FGm12 ~ s(FGm0) + Treatment + SysPres + weight, data = data)
summary(gam1.2)
gam1.3 <- gam(FGm12 ~ s(FGm0) + Treatment + SysPres, data = data)
summary(gam1.3)
gam1.4 <- gam(FGm12 ~ s(FGm0) + Treatment, data = data)
summary(gam1.4)
anova(gam1.4, gam1.0, test='F')
```
Following this strategy we see that none of these variables that where candidate to delete are significant. Deviance drops severely when dropping all these variables (from 41% to 33%). However, an ANOVA f-test indicates that the decrease in deviance of the complexer model including all variables is not significant. Our current best model is hence the simple gam1.4 model.

Before looking at tensor products and numerical-factor interactions, lets try to get insight in the gam1.4 model by using some visualizations.

This plot shows how smoothing the FGm0 variable clearly increases predictive power.

```{r}
plot(gam1.4,pages=1,residuals=TRUE, shade=TRUE, cex=2, lwd=2)
```

This gam check plot shows are model does not yet capture the data well enough (assymetrical residual vs linear pred, hist res and resp vs fitted value plots)

```{r}
par(mfrow=c(2,2))
gam.check(gam1.4)
```

Now tensor products are looked into. Three variable combinations of the variables are made. An anova test between the resulting models is also done.

```{r}
gam2.1 <- gam(FGm12 ~ s(FGm0) + Treatment + te(SysPres, weight) + te(DiaPres, height), data = data)

gam2.2 <- gam(FGm12 ~ s(FGm0) + Treatment + te(DiaPres, weight) + te(SysPres, height), data = data)

gam2.3 <- gam(FGm12 ~ s(FGm0) + Treatment + te(height, weight) + te(SysPres, DiaPres), data = data)

summary(gam2.1)
summary(gam2.2)
summary(gam2.3)
anova(gam2.1, gam2.2, gam2.3, test='F')
```

We see that model gam2.2 has a severe increase in deviance an is significant in the anova test. We thus compare it to the gam1.4 model in an anova test.

```{r}
anova(gam1.4, gam2.2, test='F')
```

GAM2.2 just barely fails to beat the null hypothesis (p value just above 5 percent). Therefore, with the increase in deviance, we consider both models our best models. 

We look at how GAM2.2 works now using plots

```{r}
plot(gam2.2,residuals=TRUE, shade=TRUE, cex=2, lwd=2)
```

```{r}
par(mfrow=c(2,2))
gam.check(gam2.2)
```

```{r}
vis.gam(gam2.2,view=c("FGm0","DiaPres"),
        theta = -40, phi = 10, r = sqrt(3), d = 1)
```
```{r}
vis.gam(gam2.2,view=c("FGm0","height"),
        theta = 0, phi = 30, r = sqrt(3), d = 1)
```

Let's now see if we can add interactions between the Treatment factor and numericals. The following models are proposed (build from our best models so far):

```{r}
gam3.1 <- gam(FGm12 ~ s(FGm0, by=Treatment) + Treatment, data=data)
gam3.2 <- gam(FGm12 ~ s(FGm0, by=Treatment) + Treatment + te(DiaPres, weight) + te(SysPres,height), data=data)
summary(gam3.1)
summary(gam3.2)
```

```{r}
anova(gam1.4, gam2.2, gam3.1, gam3.2, test='F')
```

```{r}
anova(gam2.2, gam3.1, gam3.2, test='F')
```

```{r}
anova(gam3.1, gam3.2, test='F')
```

```{r}
gam.final <- gam3.2
par(mfrow=c(2,2))
gam.check(gam.final)
```

```{r}
plot(gam.final,residuals=TRUE, shade=TRUE, cex=2, lwd=2)
```

```{r}
vis.gam(gam2.2,view=c("FGm0","DiaPres"),
        theta = -40, phi = 10, r = sqrt(3), d = 1)
```
