---
title: 'Assignment: Local Poisson Regression'
author: "AlbertMartin_AntoninRosa_LouisVanLangendonck"
date: "2023-11-10"
output:
  pdf_document: default
  html_document: default
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sm)
```


## 1: Bandwidth choice for the local Poisson regression

In order to find the optimal bandwidth through LOOCV, we must compute the LOOCV likelihood for every possible h value, and choose the one with the best odds.

In order to compute the LOOCV likelihood for a given h, we must compute the product of the individual likelihoods of each sample when the regression is fitted without it. In our case, we compute the logarithm of the likelihood, so we compute the sum of the individual log likelihoods.

The modifications performed on the h.cv.sm.binomial in order to obtain h.cv.sm.poisson were minimal, as this function simply calls loglik.CV for each possible bandwidth and chooses the best.

The modifications performed on loglik.CV in order to compute the LOOCV likelihood for a local poisson regression were the following:

 - Swap sm.binomial for sm.poisson in order to fit a local poisson regression instead. The parameters are the same, them being the sample x and y values without the current point, the current bandwidth being tested, and we want it to only be evaluated at the point not included in the fit proccess.
 
 - For the estimated lambda $\lambda_i$ at each point $x_i$, we compute the probability of obtaining the respective $y_i$ assuming it follows a poisson distribution with the estimated lambda as parameter. This is done with the dpois function, which computes the density of the given poisson distribution at $y_i$. The logarithm is taken for the likelihood at each $i$, and their mean is computed, which is the final log likelihood.

```{r}
h.cv.sm.poisson <- function(x,y,rg.h=NULL,l.h=10,method=loglik.CV){
   cv.h <- numeric(l.h)
   if (is.null(rg.h)){
      hh <- c(h.select(x,y,method="cv"),
              h.select(x,y,method="aicc"))#,hcv(x,y))
      rg.h <- range(hh)*c(1/1.1, 1.5)
   }
   i <- 0
   gr.h <- exp( seq(log(rg.h[1]), log(rg.h[2]), l=l.h))
   for (h in gr.h){
      i <- i+1
      cv.h[i] <- method(x,y,h)
   }
   return(list(h = gr.h, 
               cv.h = cv.h, 
               h.cv = gr.h[which.min(cv.h)]))
}


loglik.CV <- function(x,y,h){
  n <- length(x)
  probs <- sapply(1:n, 
      function(i,x,y,h){
         lambda<-sm.poisson(x=x[-i],y=y[-i],h=h,eval.points=x[i],display="none")$estimate
         return(dpois(y[i],lambda))
      },   x,y,h)
  
  
  return(-sum( log(probs) )/n)
}
```


## 2: Local Poisson regression for Country Development Data

First we read the data and sort it in ascending order according to the given x.

```{r}
countries<-read.csv2(file="HDI.2017.subset.csv",row.names = 1)

le.fm.r <- round(countries$le.fm)

#Sort them
perm<-order(countries$Life.expec)
Life.expec<-countries$Life.expec[perm]
le.fm.r<-le.fm.r[perm]
```

Next, we invoke the function defined in the previous exercise in order to find the ideal bandwidth. We left the determination of the range to be tested to the function, but we could have just done so manually with a list of logarithmically spaced numbers between 0.5 and 20, as we can clearly tell by looking at the data that 0.5 barely reaches a couple of consecutive entries and 20 encompasses more half of them at worst.

```{r}
cvRes<-h.cv.sm.poisson(Life.expec,le.fm.r)
cvRes$h.cv
```

With this optimal bandwidth, we fit the local poisson regression as instructed:

```{r}
sm_model<-sm.poisson(Life.expec,le.fm.r,cvRes$h.cv,eval.points=Life.expec)
```

We also wished to showcase the confidence bands, so we plotted the individuals, along with the estimation line and upper and lower confidence lines, to see how the local approximation has different cofidence depending on the x.

```{r}
plot(Life.expec,le.fm.r, col='lightgrey')
lines(Life.expec, sm_model$estimate)
lines(Life.expec, (sm_model$upper), col = 'red')
lines(Life.expec, (sm_model$lower), col = 'red')
```
