---
title: "08_IML_assignment"
author: "Louis Van Langendonck & Antonin Rosa & Albert Martín"
date: "`r format(Sys.time(), '%d/%b/%Y')`"
output: html_document
---

## 0. Data Processing

Loading in data and library. Creating a training sample choosing 700 data at random. The non-chosen data will be the test set.

```{r}
library('readxl')
library('ranger')
library('vip')
library('gridExtra')
library('fastshap')
library('mgcv')
library('grid')
library('ggplot2')
library('DALEX')
library('lime')
library('DALEXtra')
```


```{r}
set.seed(123)

concrete <- as.data.frame(read_excel("Concrete_Data.xls"))
DescVars <- names(concrete)
names(concrete) <- c("Cement","Slag","FlyAsh","Water","Superplast",
"CoarseAggr","FineAggr","Age","Strength")

train.rows <- sample(nrow(concrete), 700)

data.train <- concrete[train.rows,]
data.test <- concrete[-train.rows,]

head(data.test)
```

## 1. Fit a Random Forest

```{r}
rf.model.impurity <- ranger(formula = Strength ~ .,data = data.train, importance='impurity')
rf.model.permutation <- ranger(formula = Strength ~ .,data = data.train, importance='permutation')

pred.fun <- function(object, newdata) {
  setNames(predict(object, newdata)$predictions, row.names(newdata)) # Exact same format as predict.lm to avoid errors
}

set.seed(123)

rf.imp.vip <- vip(rf.model.impurity, num_features = 8)
rf.perm.vip <- vip(rf.model.permutation, num_features = 8)
rf.shapley <- vip(rf.model.impurity, method="shap",
                  pred_wrapper=pred.fun, num_features = 8,
                  train = data.train, newdata=data.test[,1:8])

grid.arrange(rf.imp.vip, rf.perm.vip, rf.shapley,
             ncol=2, nrow=2,
             top="Top left: Impurity. Top right: oob permutations. Bottom left: Shapley values"
            )
```


## 2. Fit a linear model and a gam model

```{r}
lm.model <- lm(Strength ~ ., data = data.train)
summary(lm.model)
```

```{r}
gam1 <- gam(Strength ~ s(Cement) + s(Slag) + s(FlyAsh) + s(Water) + s(Superplast) + s(CoarseAggr) + s(FineAggr) + s(Age), data = data.train)
summary(gam1)

gam2 <- gam(Strength ~ s(Cement) + s(Slag) + s(FlyAsh) + s(Water) + s(Superplast) + CoarseAggr + s(FineAggr) + s(Age), data = data.train)
summary(gam2)

plot(gam1)
```

```{r}
lm.shapley <- vip(lm.model, method="shap",
                  pred_wrapper=predict.lm,
                  train=data.train, # train set must be specified
                  newdata=data.test, 
                  num_features = 8,
                  exact=TRUE)

gam.shapley1 <- vip(gam1, method="shap",
                  pred_wrapper=predict.gam,
                  train=data.train, # train set must be specified
                  newdata=data.test[,1:8], 
                  num_features = 8,
                  exact=TRUE) 

gam.shapley2 <- vip(gam2, method="shap",
                  pred_wrapper=predict.gam,
                  train=data.train, # train set must be specified
                  newdata=data.test[,1:8], 
                  num_features = 8,
                  exact=TRUE) 

grid.arrange(rf.shapley, lm.shapley, gam.shapley1, 
             gam.shapley2, 
             ncol=2, nrow=2,
             top="1,1: Shapley. 1,2: Shapley lm. 2,1: Shapley gam1. 2,2: Shapley gam2"
)
```

## 3. Relevance by Ghost Variables

Note that the Relevant Ghost Variable function provided by the professor does not work on the ranger random forest model. To solve it, two adaptations are made to the function:

1) Now, a user-specified prediction function has to be given (analogous to the vip function) to fix error based on predict.ranger outputs being of a different format than 'predict.lm' and 'predict.gam' etc.

2) in checking the class model for defining the 'term.labels' variable, an extra check for 'class(model)[1]!="ranger"' is added.

```{r}
source("relev.ghost.var.R")

rf.gv <- relev.ghost.var(model=rf.model.impurity, 
                              newdata = data.test[, -9],
                              y.ts = data.test[, 9],
                              func.model.ghost.var = lm,
                              pred.wrapper = pred.fun
)

plot.relev.ghost.var(rf.gv,n1=500,ncols.plot = 4)
```

```{r}
lm.gv <- relev.ghost.var(model=lm.model, 
                              newdata = data.test[, -9],
                              y.ts = data.test[, 9],
                           func.model.ghost.var = lm,
                           pred.wrapper = predict.lm
)
plot.relev.ghost.var(lm.gv,n1=500,ncols.plot = 4)
```

```{r}
gam.gv <- relev.ghost.var(model=gam1, 
                              newdata = data.test[, -9],
                              y.ts = data.test[, 9],
                           func.model.ghost.var = lm, 
                           pred.wrapper = predict.gam
)
plot.relev.ghost.var(gam.gv,n1=500,ncols.plot = 4)
```

## 4. Global Importance Measures and Plots using the library DALEX

```{r}
explainer_rf <- explain.default(model = rf.model.impurity,  
                               data = data.test[, -9],
                               y = data.test[, 9], 
                               label = "Random Forest")
```

```{r}
Rnd_Perm <- model_parts(
  explainer_rf,
  N = NULL, # All available data are used
  B = 10   # number of permutations to be used, with B = 10 used by default
)

Rnd_Perm
plot(Rnd_Perm)
```

```{r}
aux.plot <- plot(Rnd_Perm)
dropout_loss.y <- Rnd_Perm$dropout_loss[1]
aux.I <- order(-aux.plot$data$dropout_loss.x)
rf_perm_DALEX_as_vi <- tibble::tibble(aux.plot$data[aux.I,c(2,4)])
class(rf_perm_DALEX_as_vi) <- c("vi", class(rf_perm_DALEX_as_vi))
names(rf_perm_DALEX_as_vi) <- c("Variable", "Importance")
rf_perm_DALEX_as_vi$Importance <- 
  rf_perm_DALEX_as_vi$Importance - dropout_loss.y

# Creating the ggpolt: 
rf.perm.DALEX.vip <- vip(rf_perm_DALEX_as_vi)

grid.arrange(rf.imp.vip, rf.perm.vip,
             rf.perm.DALEX.vip, ncol=2, nrow=2,
             top="Top left: Impurity. Top right: oob permutations. Bottom left: test sample permutations"
             )
```

```{r}
# ?model_profile
PDP_rf <- model_profile(
  explainer=explainer_rf,
  variables = NULL,  # All variables are used
  N = NULL, # All available data are used
  groups = NULL,
  k = NULL,
  center = TRUE,
  type = "partial" #  partial, conditional or accumulated
)

plot(PDP_rf, facet_ncol=2)
```

```{r}
CDP_rf <- model_profile(
  explainer=explainer_rf,
  variables = NULL,  # All variables are used
  N = NULL, # All available data are used
  groups = NULL,
  k = NULL,
  center = TRUE,
  type = "conditional" #  partial, conditional or accumulated
)

plot(CDP_rf, facet_ncol=2)
```

```{r}
ALE_rf <- model_profile(
  explainer=explainer_rf,
  variables = NULL,  # All variables are used
  N = NULL, # All available data are used
  groups = NULL,
  k = NULL,
  center = TRUE,
  type = "accumulated" #  partial, conditional or accumulated
)

plot(ALE_rf, facet_ncol=2)
```

## 5. Local Explainers with the library DALEX

```{r}
point.min.strength <- data.test[which.min(data.test[,9]),-9]
point.max.strength <- data.test[which.max(data.test[,9]),-9]

shap.rf.min <- predict_parts(explainer = explainer_rf,
                 new_observation = point.min.strength,
                            type = "shap")

plot(shap.rf.min)

shap.rf.max <- predict_parts(explainer = explainer_rf,
                 new_observation = point.max.strength,
                            type = "shap")

plot(shap.rf.max)
```

```{r}
bd.rf.min <- predict_parts(explainer = explainer_rf,
                 new_observation = point.min.strength,
                            type = "break_down")

plot(bd.rf.min)

bd.rf.max <- predict_parts(explainer = explainer_rf,
                 new_observation = point.max.strength,
                            type = "break_down")

plot(bd.rf.max)
```

Because the 'DALEXtra' package was built under another version of R, the following function for lime-like analysis does not work anymore. 

```{r}
#lime_rf <- predict_surrogate(explainer = explainer_rf,  DOES NOT WORK
                  #new_observation = data.test[,-9], 
                  #type = "localModel") 
                  #type = "iml") # it does not work
                  #n_features = 6, 
                  #n_permutations = 1000,
                  #type = "lime") # it does not work
```

Therefore we have the following solution. This is a special package that does LIME analysis. BLABLA

```{r}
explainer.rf.lime <- lime(data.test, rf.model.impurity)
lime.explain.min <- lime::explain(point.min.strength, explainer.rf.lime, n_features = 5)
plot_features(lime.explain.min)
lime.explain.max <- lime::explain(point.max.strength, explainer.rf.lime, n_features = 5)
plot_features(lime.explain.max)
```


```{r}
ice.rf.min <- predict_profile(explainer = explainer_rf, 
                           new_observation = point.min.strength)
plot(ice.rf.min, facet_ncol=3)

ice.rf.max <- predict_profile(explainer = explainer_rf, 
                           new_observation = point.max.strength)
plot(ice.rf.max, facet_ncol=3)
```

```{r}
mp_rf <- model_profile(explainer = explainer_rf,
  variables = "Age",
  N = NULL, #All observations in test_set
  type = "partial"
)

?model_profile()

plot(mp_rf, geom = "profiles") +  
  ggtitle("Ceteris-paribus and partial-dependence profiles for Age") 
```
