---
title: "Density estimation. GMM. DBSCAN (Assignment)"
author: "Louis Van Langendonck & Antonin Rosa"
date: "`r format(Sys.time(), '%d/%b/%Y')`"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
load("BikeDay.Rdata")
X <- as.matrix(day[day$yr==1,c(10,14)])
pairs(X)
```

## 1. model based clustering assuming a Gaussian Mixture Model

Using the G parameter in the `Mclust` function, options for K are set between 2 and 6. It automatically selects the model with the highest BIC score. Moreover, the BIC plot Via the `modelNames` parameter, mixture components are allowed variation in volume, shape, and orientation. The 'BIC'-plot shows that K = 3 scores the best and is thus the choice of K the model automatically continues with. The classification, uncertainty and density plots use this value of K

```{r}
library(mclust)
GMM <- Mclust(X,G=2:6,modelNames="VVV")
plot(GMM, "BIC")
plot(GMM, "classification")
plot(GMM, "uncertainty")
plot(GMM, "density")
```

## 2. Compare the previous density plot with the non-parametric density estimation of (temp,casual).

Comparing these plots, it becomes clear both show a similar profile. However the GMM model (left) yields a clear distinction in three gaussian shapes (as typically predicted in this type of model), while the non-param model (right) shows more arbitrary outlines.

```{r}
library(sm)
plot(GMM,"density",main="GMM")
a <- 0.25
h_prop <- a*c(sd(X[,"temp"]), sd(X[,"casual"]))
sm.density(X,h_prop,display="slice",main="Non-param Denisty Estimator")
```

## 3. Non-param density estimation of each of the clusters

```{r}
clust.ind <- GMM$classification
plot(X,col=clust.ind)
for (j in 1:3){
  cl.j <- (clust.ind==j)
  sm.density(X[cl.j,],h=0.4*c(sd(X[cl.j,"temp"]),sd(X[cl.j,"casual"])),display="slice",props=c(75),col=j, cex=4, add=TRUE)
}
```

## 4. Component merging

```{r}
library(fpc)
GMMbic <- mclustBIC(X,G=3,modelnames='VVV')
GMMmerge <- fpc::mergenormals(X,summary(GMMbic,X),method = "bhat")
GMMmerge.ind <- GMMmerge$clustering
plot(X,col=GMMmerge.ind)
```

## 5. Non-param density estimation of each of the new, merged clusters

```{r}
plot(X,col=GMMmerge.ind)
for (j in 1:2){
  cl.j <- (GMMmerge.ind==j)
  sm.density(X[cl.j,],h=0.4*c(sd(X[cl.j,"temp"]),sd(X[cl.j,"casual"])),display="slice",props=c(75),col=j, cex=4, add=TRUE)
}
```


## 6. DBScan

```{r}
library(dbscan)
library(cluster)

Xs <- scale(X)

eps_values <- c(0.25, 0.5)
minPts_values <- c(10, 15, 20)
results <- data.frame()

best_silhouette <- -Inf
best_eps <- NULL
best_minPts <- NULL

for (eps in eps_values) {
  for (minPts in minPts_values) {
    dbscan_result <- dbscan(Xs, eps = eps, minPts = minPts)
    
    # Vérifiez s'il y a plus d'un cluster
    print(max(dbscan_result$cluster))
    if (max(dbscan_result$cluster) > 1) {
      silhouette_avg <- silhouette(dbscan_result$cluster, dist(Xs))
      avg_silhouette <- mean(silhouette_avg[, "sil_width"])
      
      results <- rbind(results, data.frame(eps = eps, minPts = minPts, silhouette = avg_silhouette))
    
      if (avg_silhouette > best_silhouette) {
        best_silhouette <- avg_silhouette
        best_eps <- eps
        best_minPts <- minPts
      }
    }
  }
}

cat("Best combination of parameters:\n")
cat("Epsilon:", best_eps, "\n")
cat("MinPts:", best_minPts, "\n")

dbscan_result <- dbscan(Xs, eps = best_eps, minPts = best_minPts)
plot(Xs, col = dbscan_result$cluster, pch = 19, main = "DBSCAN Clustering")

contingency_table <- table(dbscan = dbscan_result$cluster, GMM = GMMmerge.ind)

print(contingency_table)

```

Even though DBSCAN Cluster 0 (representing noise or unclustered data) is attributed to two distinct clusters by GMM (GMM Cluster 0 and GMM Cluster 1), the total number of points in this cluster is relatively low (7 points in total).

Overall, these results indicate a strong similarity between the clustering results of DBSCAN and GMM for most observations. Ambiguity is primarily found in DBSCAN Cluster 0, representing noise or unclustered data. This similarity may suggest that both methods are uncovering similar data structures, at least for the data that is meaningfully clustered.