---
title: "Smoothing and regression splines"
author: "Louis Van Langendonck & Antonin Rosa & Albert Martín"
date: "`r format(Sys.time(), '%d/%b/%Y')`"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

### Question 1

#### Question a

We load in the data and apply the smooth.spline function to find smoothing parameters. To use the generalized cross validation parameter, we have to specify cv = false (is true it is ordinary leave-one-out cross-validation)

```{r}
load("bikes.Washington.Rdata")

spline_fit_cv <- smooth.spline(bikes$instant, bikes$cnt, cv = FALSE, all.knots = FALSE)

# Extract information from the fit with cross-validation
lambda_cv <- spline_fit_cv$spar
lambda_cv
```

The value of the chosen penalty parameter $\lambda$ is 0.24802, determined by generalized cross-validation.

#### Question b

```{r}
df_cv <- spline_fit_cv$df
df_cv
```

The value of df determined by cross-validation is 93.34091.

#### Question c

```{r}
knots_cv <- spline_fit_cv$fit$knot
print(length(knots_cv))
```

During the smoothing and the regression 140 knots have been used.

#### Question d

```{r}
plot(bikes$instant, bikes$cnt, main = "Scatterplot and Estimated Regression Function (CV)",
     xlab = "instant", ylab = "cnt", col = "blue")
lines(spline_fit_cv, col = "red")
```

### Question 2

Creating the new cnt.5000 binary variable that sets 1 when bike counts are higher or equal to 5000.

```{r}
cnt.5000 <- rep(0, length(bikes$cnt))
cnt.5000[which(bikes$cnt>=5000)] <- 1
```

Loading in the R function that implements the IRWLS algorithm.

```{r}
source("IRWLS_logistic_regression.R")
```

### Question a

```{r}
spline.irwls.bike <- logistic.IRWLS.splines(x=bikes$temp, y=cnt.5000, df=6)

plot(bikes$temp,cnt.5000)
lines(bikes$temp,spline.irwls.bike$fitted.values,col=3)
temp.int.start <- min(bikes$temp[which(spline.irwls.bike$fitted.values>0.5)])
temp.int.end <- max(bikes$temp[which(spline.irwls.bike$fitted.values>0.5)])
abline(h=0.5,col=2)
abline(v=temp.int.start,col=4)
abline(v=temp.int.end,col=4)
```

### Question b

```{r}
#install.packages("caret")
require(caret)
flds <- createFolds(cnt.5000, k = 5, list = TRUE)

df.v = 3:15

loglik.dfs <- c()
for (df.proposal in df.v){
  loglik.folds <- c()
  for (i in (1:length(flds))){
    spline.irwls.bike.ifold <-  logistic.IRWLS.splines(x=bikes$temp[-unlist(flds[i])],y=cnt.5000[-unlist(flds[i])],x.new=bikes$temp[unlist(flds[i])],df=df.proposal)
    
    # log_likelihood_binary = sum_i[(yi*log(p(xi)) + ((1-yi)*log(1-p(xi)))]
    loglik_term1 <- cnt.5000[unlist(flds[i])]*log(spline.irwls.bike.ifold$predicted.values)
    loglik_term2 <- (1-cnt.5000[unlist(flds[i])])*log(1-(spline.irwls.bike.ifold$predicted.values))
    loglik.folds <- c(loglik.folds,sum(loglik_term1 + loglik_term2))
  }
  loglik.dfs <- c(loglik.dfs, sum(loglik.folds))
}
plot(df.v, loglik.dfs)
abline(v=df.v[which.max(loglik.dfs)],col=4)
```

```{r}
spline.irwls.bike.loglike <- logistic.IRWLS.splines(x=bikes$temp, y=cnt.5000, df=4)
plot(bikes$temp,cnt.5000)
lines(bikes$temp,spline.irwls.bike.loglike$fitted.values,col=3)
temp.int.start <- min(bikes$temp[which(spline.irwls.bike.loglike$fitted.values>0.5)])
temp.int.end <- max(bikes$temp[which(spline.irwls.bike.loglike$fitted.values>0.5)])
abline(h=0.5,col=2)
abline(v=temp.int.start,col=4)
abline(v=temp.int.end,col=4)
```