---
title: "Density estimation: Bandwidth choice by leave-one-out maximum likelihood"
author: "Louis Van Langendonck"
date: "2023-09-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Histogram

#### 1.

Consider the kernel density estimator at $x_i$ using all the observations : $$
\hat{f}_{hist}(x_i) = \frac{1}{nb} \sum_{k=1}^n I_{B_{j\left(x_k\right)}}(x_i) $$. Which can be rewritten as: $$
\hat{f}_{hist}(x_i) = \frac{1}{nb} \left(1 + \sum_{k \neq i}^n I_{B_{j\left(x_k\right)}}(x_i) \right) $$. Now consider the expression for the leave-one-out likelihood of observation $x_i$ (so computed by all other elements in the sample): $$ \hat{f}_{hist, (-i)}(x_i) = \frac{1}{(n-1)b} \sum_{k \neq i}^n I_{B_{j\left(x_k\right)}}(x_i) $$, which can then be rewritten using the previously derived formula for $\hat{f}_{hist}(x_i)$ to the final expression: $$ \hat{f}_{hist, (-i)}(x_i) = \left(\frac{n}{n-1}\right) \left( \hat{f}_{hist}(x_i) - \frac{1}{nb} \right)$$.

#### 2-3.

```{r}
cdrate.df <- read.table("cdrate.dat")
x <- cdrate.df[,1]

A <- min(x)-.05*diff(range(x)) 
Z <- max(x)+.05*diff(range(x))
nbr <- 7

hx <- hist(x,breaks=seq(A,Z,length=nbr+1),freq=F)
binwidth <- hx$breaks[2]-hx$breaks[1]
hx_f <- stepfun(hx$breaks,c(0,hx$density,0))
y_hist <- hx_f(x)
points(x,y_hist,col="red", pch=1)

y_loo_hist <- (y_hist-(1/(length(x)*binwidth)))*(length(x)/(length(x)-1))
points(x,y_loo_hist,col="blue", pch=1)

par(xpd=TRUE)
legend( x="topleft", 
        legend=c(expression("f_hist"),"f_hist l-o-o"),
        col=c("red","blue"), pch=c(1,1))
```

#### 4.

The sample log-likelihood function is calculated using $l_{C V}(b)=\sum_{i=1}^n \log \hat{f}_{hist,(-i)}\left(x_i\right)$.

```{r}
loo_log_likelihood <- sum(log(y_loo_hist))
```

If we apply this to our data and estimators (where the binsize b is equal to `r binwidth`), the result is $l_{C V}(b)=$`r loo_log_likelihood`.

#### 5.

```{r}
nbr_seq <- seq(1,15)
nbr_loo_log_likelihood <- c()
for (nbr in nbr_seq) {
  hx <- hist(x,breaks=seq(A,Z,length=nbr+1),plot=F)
  binwidth <- hx$breaks[2]-hx$breaks[1]
  hx_f <- stepfun(hx$breaks,c(0,hx$density,0))
  y_hist <- hx_f(x)
  y_loo_hist <- (y_hist-(1/(length(x)*binwidth)))*(length(x)/(length(x)-1))
  if (any(y_loo_hist==0)) {
    nbr_loo_log_likelihood <- c(nbr_loo_log_likelihood,-Inf)
  }
  else {
    nbr_loo_log_likelihood <- c(nbr_loo_log_likelihood,sum(log(y_loo_hist)))
  }
}
```

Some of these log likelihoods will give -inf values. This happens when an observation has a leave-one-out likelihood of zero of which the log cannot be calculated.

```{r}
plot(nbr_seq,nbr_loo_log_likelihood, pch=16, xlab='amount of bins', ylab = 'log likelihood (leave-one-out)')
```

From the plot is becomes clear that, based on the leave-one-out max-likelihood metric, 5 bins is the optimal amount of bins (from the proposed sequence). The corresponding histogram is plotted below

```{r}
nbr_opt <- 5
hx <- hist(x,breaks=seq(A,Z,length=nbr_opt+1),freq=F)
```

#### 6.

Very similarly as the previous experiment, optimal kernel density estimator bandwidth is chosen. This time by directly varying bin width b instead of the amount of bins.

```{r}
b_seq <- seq((Z-A)/15,(Z-A)/1,length=30)
b_loo_log_likelihood <- c()
for (b in b_seq) {
  hx <- hist(x,breaks=seq(A,Z+b,by=b), plot=F)
  hx_f <- stepfun(hx$breaks,c(0,hx$density,0))
  y_hist <- hx_f(x)
  y_loo_hist <- (y_hist-(1/(length(x)*b)))*(length(x)/(length(x)-1))
  if (any(y_loo_hist==0)) {
    b_loo_log_likelihood <- c(b_loo_log_likelihood,-Inf)
  }
  else {
    b_loo_log_likelihood <- c(b_loo_log_likelihood,sum(log(y_loo_hist)))
  }
}
plot(b_seq,b_loo_log_likelihood, pch=16, xlab='Bin Width', ylab = 'log likelihood (leave-one-out)')
```

```{r}
b_optimal <- b_seq[which.max(b_loo_log_likelihood)]
```

From the bin-width vs. log-likelihood plot it becomes clear that the optimal value for the bin width (of the proposed sequence) is `r b_optimal`. The corresponding histogram is as follows:

```{r}
hx <- hist(x,breaks=seq(A,Z+b,by=b_optimal), plot=F)
plot(hx,freq = FALSE)
```

```{r mixt.normal, echo=FALSE}
# graph.mixt
# Input:
#    k: number mixture components
#    mu: vector of length k with the mean values of the k normals
#    sigma: vector of length k with the st.dev. values of the k normals
#    alpha: vector of length k with the weights of each normal
#    graphic: logical value indicating if the mixture density must be plotted
#    ...: Other parameters passed to plot()
#
# Output:
#    L, U: extremes of the interval where the mixture density is plotted
#    x: points at which the mixture density is evaluated 
#    fx: value of the mixture density at x
#
graph.mixt<-
function(k=1, mu=seq(-2*(k-1),2*(k-1),length=k), sigma=seq(1,1,length=k), alpha=seq(1/k,1/k,length=k), graphic=TRUE,...)
{
   L<-min(mu-3*sigma)
   U<-max(mu+3*sigma)
         
   x<- seq(from=L,to=U,length=200)
   fx<- 0*x
   Salpha<-sum(alpha)
   for(i in 1:k){
   	p<-alpha[i]/Salpha
#   	fx <- fx + p*exp(-.5*((x-mu[i])/sigma[i])^2)/(sqrt(2*pi)*sigma[i])
   	fx <- fx + p*dnorm(x,mu[i],sigma[i])
   }
   if (graphic){
      plot(x,fx,type="l",...)
   }
   return(list(L = L, U = U, x = x, fx = fx))
}

# sim.mixt
# Input:
#    n: number of simulated data
#    k: number mixture components
#    mu: vector of length k with the mean values of the k normals
#    sigma: vector of length k with the st.dev. values of the k normals
#    alpha: vector of length k with the weights of each normal
#    graphic: logical value indicating if the mixture density and the 
#              histogram of the simulated data must be plotted
#    ...: Other parameters passed to plot()
#
# Output:
#    x: simulated data
#
# Requires: 
#    graph.mixt
sim.mixt <- function(n=1,k=1, 
         mu=seq(-2*(k-1),2*(k-1),length=k), 
         sigma=seq(1,1,length=k), 
         alpha=seq(1/k,1/k,length=k), graphic=FALSE,...)
{
   csa<-cumsum(alpha)
   x<-runif(n)
      
   for (i in 1:n){
      comp<-sum(csa<=x[i])+1
      x[i]<-rnorm(1,mu[comp],sigma[comp])
   }
   if(graphic) {
      out<-graph.mixt(k, mu, sigma, alpha, gr=FALSE)
      hist(x,freq = FALSE,
           ylim=c(0,max(c(max(out$fx),max(hist(x,plot=FALSE)$density)))))
      lines(out$x,out$fx,lty=1,lwd=2)
   }   
   return(x)
}
```

#### 7.

Generate $n=100$ data from $$
f(x) = (3/4)N(x; m = 0, s = 1) +(1/4) N(x; m = 3/2, s = 1/3)$$

```{r}
set.seed(123)
n <- 100
mu <- c(0,3/2)
sigma <- c(1,1/3)
alpha <- c(3/4,1/4)
x <- sim.mixt(n=n, k=2, mu=mu, sigma=sigma, alpha=alpha, gr=T)
points(x,0*x,pch="|")

sigma.mixt <- 1.095287
b_scott <- 3.49 * sigma.mixt * length(x)^(-1/3) 
```

First we use the Scott's formula to find a proposed value of b = `r b_scott` (calculations for this value can be found in `density_estimation.Rmd`). The corresponding histogram is as follows:

```{r}
A <- min(x)-.05*diff(range(x)) 
Z <- max(x)+.05*diff(range(x))

hx <- hist(x,breaks=seq(A,Z+b,by=b_scott), plot=F)
plot(hx,freq = FALSE)
```

Now the Max-likelihood method is used. To consider a varied amount of proposals for b, the sequence of proposed b-values to select from is varied from 0.05 to 1.75 in steps of 0.01.

```{r}
b_seq = seq(0.05, 1.75, by=0.01)
b_loo_log_likelihood <- c()

for (b in b_seq) {
  hx <- hist(x,breaks=seq(A,Z+b,by=b), plot=F)
  hx_f <- stepfun(hx$breaks,c(0,hx$density,0))
  y_hist <- hx_f(x)
  y_loo_hist <- (y_hist-(1/(length(x)*b)))*(length(x)/(length(x)-1))
  if (any(y_loo_hist<=0)) {
    b_loo_log_likelihood <- c(b_loo_log_likelihood,-Inf)
  }
  else {
    b_loo_log_likelihood <- c(b_loo_log_likelihood,sum(log(y_loo_hist)))
  }
}
plot(b_seq,b_loo_log_likelihood, pch=16, xlab='Bin Width', ylab = 'log likelihood (leave-one-out)')
```

```{r}
b_ml <- b_seq[which.max(b_loo_log_likelihood)]
```

From this experiment, the leave-one-out cross-validation max log likelihood method selects b = `r`b_ml` as the optimal value for b (out of the proposed values for b). The corresponding histogram can be found below.

```{r}
hx <- hist(x,breaks=seq(A,Z+b,by=b_ml), plot=F)
plot(hx,freq = FALSE)
```

The two proposed values for b are substantially different. The corresponding histograms/kernel density estimators clearly show the difference the choice of b makes: the corresponding patterns seem to give quite a different view on the data. Using Scott's formula, the mixture of two distributions (here gaussians) is not at all clearly visible, while the max-likelihood method's bin width seems to highlight more the presence of two distinct patterns.

#### 8.

In this exercise, the *gaussian* kernel is used and its bandwidth found similarly to the bin width in the previous exercises. A wide range of bandwidth values are proposed (from 0.1 to 1 with steps of 0.01) and the a plot of each leave-one-out log-likelihood provided.

```{r}
h_seq <- seq(0.1,1, by=0.01)
h_loo_log_likelihood <- c()

for (h in h_seq) {
  kx <- density(x,bw=h,kernel='gaussian')
  kx_f <- approxfun(x=kx$x, y=kx$y, method='linear', rule=2)
  y_density <- kx_f(x)
  y_loo_density <- (y_density-(1/(sqrt(2*pi)*length(x)*h)))*(length(x)/(length(x)-1))
  
    h_loo_log_likelihood <- c(h_loo_log_likelihood,sum(log(y_loo_density)))
}

plot(h_seq,h_loo_log_likelihood, pch=16, xlab='Kernel Bandwidth', ylab = 'log likelihood (leave-one-out)')

h_ml <- h_seq[which.max(h_loo_log_likelihood)]
```

From this experiment, it becomes clear that the optimal max likelihood value for h is `r h_ml`. The corresponding density function is plotted below.

```{r}
kx <- density(x,bw=h_ml,kernel='gaussian')
kx_f <- approxfun(x=kx$x, y=kx$y, method='linear', rule=2)
plot(kx$x, kx$y, xlab='x', ylab = 'density', main = 'Gaussian Kernel Density Estimation')
```
